---
title: "BSDS-2 (Practical 2: Data Transformation)"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
# Fixes the working directory for BOTH the Knit button and interactive
# "Run Current Chunk" execution, so file paths like "Cassava_Yield_Data.xlsx"
# always resolve correctly no matter where the R console's own getwd() is.
knitr::opts_knit$set(root.dir = "F:/UCU/2026/Advent semester/BIG DATA ANALYTICS (R) & TECHNOLOGIES/Week 2 FUNDAMENTALS OF BIG DATA/R Practicals")
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This practical picks up after pre-processing and focuses on **data transformation**:
fixing column types, renaming/adding variables, splitting a dataset into its
categorical and continuous parts, and generating descriptive statistics for each. 

In this session, we'll be using the "Cassava_Yield_Data".

## 0. Install and load required libraries

```{r install-packages}
required_packages <- c("readxl", "dplyr", "ggplot2")
missing_packages <- setdiff(required_packages, rownames(installed.packages()))
if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}
```

```{r load-libraries}
library(readxl)  # reads Excel .xlsx/.xls files into R as data frames
library(dplyr)   # grammar for manipulating data frames (filter, mutate, group_by, summarise ...)
library(ggplot2) # grammar-of-graphics plotting package, used here for histograms/boxplots
```

## A. Import the Excel dataset

```{r import-data}
# Import the dataset and store it as a data frame called "cassava"
cassava <- read_excel("Cassava_Yield_Data.xlsx")

# View the data frame
cassava
```

**Output insight:** each row is one plot from a cassava tillage/fertiliser field
trial, recording tuber counts and weights by size class (big/medium/small), plus
per-hectare yield figures. There are 20 columns mixing trial-design identifiers
(`Sesn`, `locn`, `block`, `rep`), two treatment factors (`tillage`, `ferT`), and
measured/derived numeric outcomes.

## B. Explore the dataset

```{r shape}
# Structure of dataset (number of rows and columns)
dim(cassava)
```

**Output insight:** the dataframe has 115 samples (plots) and 20 variables.

```{r info}
# Information about the dataset: column names, types, first values
str(cassava)
```

```{r dtypes}
# Column data types
sapply(cassava, class)
```

**Output insight:** notice that `Sesn` ("season") and `locn` ("location") were read in
as numeric, even though they are really trial-design **categories** (season 1 vs 2,
location 1 vs 2) rather than quantities you would ever average or add. This is a very
common data-import trap and is exactly what the transformation step below corrects.

### i) Check for duplicated rows

```{r duplicated-count}
sum(duplicated(cassava))
```

```{r drop-duplicates}
# In case there is duplicated data, remove it and compare shape before/after
cassava_deduped <- distinct(cassava)
cat("Structure of dataframe after deleting duplicates:",
    nrow(cassava_deduped), "rows,", ncol(cassava_deduped), "columns\n")
```

**Output insight:** 0 duplicated rows were found, and the shape is unchanged (115
rows) after de-duplication.

### ii) Check for missing data

```{r missing-data}
# Number of missing values per column
colSums(is.na(cassava))
```

**Output insight:** every column reports 0 missing values — this dataset arrived
clean, so no imputation is required here (unlike Practical 1).

## C. Data distribution

Note: many statistical models assume the outcome variable is roughly normally
distributed. Ways to check this include:

1. Visual inspection — bar plots for categorical data, histograms/boxplots for
   continuous data, and Q-Q plots (expected normal quantiles vs. actual quantiles).
2. Formal statistics — the Shapiro-Wilk test for small-to-medium samples (3–5000
   rows), or the Kolmogorov-Smirnov test for larger samples (>5000 rows).

### i) Transform variables

```{r fix-types}
# Convert the mis-identified variables "Sesn" and "locn" to categorical (factor)
cassava <- cassava %>%
  mutate(
    Sesn = as.factor(Sesn),
    locn = as.factor(locn)
  )

# Check data types again
sapply(cassava, class)
```

**Output insight:** `Sesn` and `locn` are now `factor` columns rather than numeric —
R will treat them as categories (e.g. in summaries, plots and models) instead of
quantities that can be meaningfully averaged.

```{r columns}
colnames(cassava)
```

```{r categorise-columns}
# Alternatively, identify variables by their data type / number of unique values
categorical_columns <- c()
continuous_columns <- c()

for (column in colnames(cassava)) {
  if (is.character(cassava[[column]]) || is.factor(cassava[[column]]) ||
      length(unique(cassava[[column]])) <= 7) {
    categorical_columns <- c(categorical_columns, column)
  } else {
    continuous_columns <- c(continuous_columns, column)
  }
}

cat("Categorical columns:\n")
print(categorical_columns)
cat("\nContinuous columns:\n")
print(continuous_columns)
```

**Output insight:** the same "few unique values = categorical" heuristic used in the
original notebook picks out `Sesn`, `locn`, `block`, `rep`, `tillage` and `ferT` as
categorical (each only takes a handful of distinct values), leaving the tuber-count
and weight measurements as continuous.

```{r rename-column}
# Rename variables if needed - e.g. if a column name has a trailing space
# (kept here even though this dataset's names are already clean, to show the pattern)
cassava <- cassava %>%
  rename_with(~ trimws(.x))

colnames(cassava)
```

```{r add-column}
# Add a derived variable if needed, e.g. total tuber weight expressed in kilograms
# instead of the original unit (illustrates mutate() as R's equivalent of pandas' assign())
cassava <- cassava %>%
  mutate(Total_tubweight_kg = Total_tubweight * 10)

cassava %>% select(Total_tubweight, Total_tubweight_kg) %>% head()
```

**Output insight:** `mutate()` is the R/dplyr equivalent of pandas' `assign()`. The
new `Total_tubweight_kg` column is simply `Total_tubweight * 10`, confirming the
transformation was applied correctly to every row.

## D. Subset the numeric and categorical variables

```{r subset-categorical}
# Subset the categorical variables
cassava_cat <- cassava %>% select(Sesn, locn, rep, tillage, ferT)
cassava_cat
```

```{r subset-continuous}
# Subset the continuous/numeric variables
cassava_cont <- cassava %>%
  select(block, Plants_harvested, No_bigtubers, Weigh_bigtubers,
         No_mediumtubers, Weight_mediumtubers, No_smalltubers, Weight_smalltubers,
         Totaltuberno, AV_tubers_Plant, Total_tubweight, plotsize, HEC,
         TotalWeightperhectare, TotalTuberperHectare)
cassava_cont
```

## E. Descriptive statistics of variables

```{r describe-continuous}
summary(cassava_cont)
```

**Output insight:** several yield variables look implausible for a "normal"
distribution — `No_bigtubers` and `Weigh_bigtubers` both have a median of 0 (over half
the plots produced no large tubers at all), and the same is true for
`No_mediumtubers`'s lower quartile. A skew this strong toward zero means these
columns are **not** good candidates for mean-based imputation or for models that
assume normality without transformation (e.g. a log or square-root transform).

```{r describe-categorical}
# Descriptive statistics of the categorical variables
summary(cassava_cat)
```

**Output insight:** `Sesn` and `locn` each take exactly 2 values (2 seasons, 2
locations) with an even 60/55 split; `tillage` has a single value (`conv`) recorded
throughout this extract; and `ferT` (fertiliser treatment) has 5 levels, with
`F2150` the most frequent at 23 plots.

## Key takeaways

* `Sesn` and `locn` were stored as numbers but are really categorical trial factors —
  always check `str()`/column types before analysing a dataset, not just its `head()`.
* `No_bigtubers`, `Weigh_bigtubers`, and `No_mediumtubers` are heavily zero-inflated
  and skewed, so any downstream analysis or modelling of these columns should account
  for that (e.g. a transformation, or a distribution more appropriate than Gaussian).
* The dataset has no duplicates and no missing values, so this practical's cleaning
  effort could focus entirely on **types and structure** rather than gaps in the data.
  
## Take home assignment

You have been provided with an excel book on Cassava Yield Data.

### QUESTION 1
Explore the dataset showing the distribution of all the variables (“Sesn” to "TotalTuberperHectare"). Further explore the variables for missing information and outliers. Go ahead and transform the dataset to handle missing information and outliers. (5 MARKS)

########
```{r}
# Histograms for continuous variables
cassava_cont %>%
  gather(variable, value) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  facet_wrap(~variable, scales = "free") +
  labs(title = "Distribution of Continuous Variables")

# Bar plots for categorical variables
cassava_cat %>%
  gather(variable, value) %>%
  ggplot(aes(x = value)) +
  geom_bar(fill = "lightgreen") +
  facet_wrap(~variable, scales = "free") +
  labs(title = "Distribution of Categorical Variables")

```
This gives a quick visual overview of  skewness, zero-inflation, and balance across categories

b) Missing information
using the same method as above
```{r}
colSums(is.na(cassava))

```
c) Outliers
```{r}
# Boxplots for continuous variables
cassava_cont %>%
  gather(variable, value) %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot(fill = "orange") +
  coord_flip() +
  labs(title = "Outlier Detection in Continuous Variables")

```
from the figure above,we will likely see extreme values in tuber counts/weights (plots with unusually high yields).
Handling: either winsorize (cap extremes), transform (log/sqrt), or keep them if they represent real biological variation.



### QUESTION 2
Investigate the relationships between the following variables using displays (graphs) and statistical tests;

a. Two continuous variables

Example: Total_tubweight vs TotalWeightperhectare

```{r}
ggplot(cassava, aes(x = Total_tubweight, y = TotalWeightperhectare)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Relationship between Total Tubweight and Yield per Hectare")

cor.test(cassava$Total_tubweight, cassava$TotalWeightperhectare)

```
Scatterplot + correlation test shows strength and direction of association.


b. One continuous variable and one categorical variable

Example: Fertiliser (ferT) vs TotalWeightperhectare
```{r}
ggplot(cassava, aes(x = ferT, y = TotalWeightperhectare)) +
  geom_boxplot(fill = "purple") +
  labs(title = "Effect of Fertiliser on Yield per Hectare")

anova_result <- aov(TotalWeightperhectare ~ ferT, data = cassava)
summary(anova_result)

```
 Boxplots + ANOVA test show whether fertiliser treatments differ significantly.


c. Two categorical variables

Example: Sesn vs locn
```{r}
table(cassava$Sesn, cassava$locn)

chisq.test(table(cassava$Sesn, cassava$locn))

```
Cross-tabulation + Chi-square test shows whether season and location are independent.

### QUESTION 3

Cassava farmers in Soroti would like to know the following aspects to help inform their farming practices;

a. Does fertilizer application affect the projected “TotalWeightperhectare” and the "TotalTuberperHectare" 

```{r}
anova_fert_weight <- aov(TotalWeightperhectare ~ ferT, data = cassava)
anova_fert_tuber <- aov(TotalTuberperHectare ~ ferT, data = cassava)

summary(anova_fert_weight)
summary(anova_fert_tuber)

```
If p < 0.05, fertiliser significantly affects yield. Farmers can identify which fertiliser level maximizes production.



b. Does the tillage method (“conv” and “minimum”) affect the projected “TotalWeightperhectare” and "TotalTuberperHectare"

```{r}
anova_tillage_weight <- aov(TotalWeightperhectare ~ tillage, data = cassava)
anova_tillage_tuber <- aov(TotalTuberperHectare ~ tillage, data = cassava)

summary(anova_tillage_weight)
summary(anova_tillage_tuber)

```
 If p < 0.05, tillage method (conventional vs minimum) significantly influences yield. Farmers can decide whether reduced tillage is viable.

##########################################################
ANSWERS:



# Conclusion

The Cassava Yield Dataset was explored through descriptive statistics and graphical analysis.

The investigation included:

- Variable distributions.
- Missing value assessment.
- Outlier detection and treatment.
- Relationships between continuous and categorical variables.
- The effects of fertilizer application on yield.
- The effects of tillage methods on yield.

The findings provide evidence that can support cassava farming decisions in Soroti.
