---
title: "Cassava Yield Analysis
output: html_notebook
---

# 1. Introduction

This notebook analyses cassava yield data to investigate how tillage and fertilizer treatments affect cassava production.

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Ctrl+Shift+Enter*. 

```{r}
plot(cars)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
```{r}
packages <- c("readxl", "dplyr", "ggplot2", "tidyr")

installed <- rownames(installed.packages())

for (p in packages) {
  if (!(p %in% installed)) {
    install.packages(p)
  }
}
```


```{r}
library(readxl
```
```{r}
df <- read_excel(
  "Cassava_Yield_Data.xlsx",
  sheet = "Cassava Data"
)
```

```{r}
head(df)
```
```{r}
dim(df)
```
```{r}
names(df)
```
```{r}
df$block
```
```{r}
str(df)
```
```{r}
summary(df)
```

